---
title: "TextureBasedSegmentation"
author: "Shruti"
date: "2023-09-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Practicing code to implement simple GLCM texture based segmentation

# load necessary libraries

```{r}
# package for vector and raster manipulation
library(terra)
# package for calculating texture metrics
library(GLCMTextures)
# package for segmentation
library(supercells)
library(sf)
library(dplyr)
```

## read the TIFF image and plot it


```{r}

# path to UAV
in.dir = "X:/delta_sav/raster/UAVimages/BlacklockComp_exp_32cm_sub_mskd.img"

# read the file
blklock = terra::rast(in.dir)
# print basic information about the file
blklock
```

```{r}
# plot band 4
plot(blklock[[4]])
```

```{r}
# plot a true color image
plotRGB(blklock, r = 4, g = 2, b = 3)

```

## Run the GLCM matrix to calculate a bunch of texture measures based on Haralick et al. 1989

https://cran.r-project.org/web/packages/GLCMTextures/vignettes/README.html

good website for explanation of the GLCMTextures package and how to use it 
to calculated texture metrics

Citation for this package:
Ilich, Alexander R. 2020. “GLCMTextures.”, https://doi.org/10.5281/zenodo.4310186. https://github.com/ailich/GLCMTextures

Alternative workflow in this ppt: https://r-spatialecology.github.io/ialena-2020/#1
Citation for Fragstats in R: https://onlinelibrary.wiley.com/doi/10.1111/ecog.04617


```{r}

# run the glcm matrix to calculate all texture measures
# only run on the green band (can try with NIR too)
blklock.glcm <- glcm_textures(blklock[[2]], w = c(5,5), n_levels = 16, quantization = "equal range", 
                              shift = list(c(1, 0), c(1, 1), c(0, 1), c(-1, 1)))

# where w = the window size over which the texture is calculated
# n_levels = number of quantization levels to reduce the band to (has a huge impact on computation time)
# quantization = method of reducing gray levels - in this case, equalizing the histogram (equal prob)
# shift = what direction the texture is calculated in - default is all four directions as shown here in this list

plot(blklock.glcm)
```


```{r}
plot(blklock.glcm$glcm_entropy)

```

## segment the image based on GLCM measures and display it

Based on methods described in this citation:
Achanta, Radhakrishna, Appu Shaji, Kevin Smith, Aurelien Lucchi, Pascal Fua, and Sabine Süsstrunk. 2012. “SLIC Superpixels Compared to State-of-the-Art Superpixel Methods.” IEEE Transactions on Pattern Analysis and Machine Intelligence 34 (11): 2274–82.

Package: supercells
Not yet figured out how to combine cells of same color

```{r}

# first var is the image to be segmented - I only gave it the entropy band
# k is the number of supercells required
# compactness is the standard shape of each segment. Higher the number, more regular the shape
# iter is number of iterations to be performed; default is 10
seg.tex = supercells(blklock.glcm$glcm_entropy, k = 1000, compactness = 1, iter = 20)

# plot the NIR band of the UAV image
plot(blklock[[2]])
# plot the segmented layer
plot(seg.tex)

```


## export the polygons; combine them with attribute data; merge if possible

```{r}
# checking if I can output the polygons
# none of the attributes are associated with the geometry
try.vec = st_sf(st_geometry(seg.tex$geometry))

# hence writing out the attribute table to join to polygons
try.df = data.frame(seg.tex) %>% mutate(phrag = ifelse(glcm_entropy > 2.5, 1, 
                                                ifelse(glcm_entropy == 0, 0,
                                                ifelse(glcm_entropy < 0.52, 2, 3))))

# convert the data frame to a points vector
cen.poly <- st_as_sf(try.df, coords = c("x", "y"), crs = 32610)
#plot(cen.poly)

# join the polygons to their attributes
seg.poly <- st_join(try.vec, cen.poly)
plot(seg.poly)

# dissolve into just two classes 
seg.cls <- seg.poly %>% group_by(phrag) %>% summarize()
plot(seg.cls)

# create the output vector file name
#outname = paste0(dirname(in.dir), "/blklock_segmented_v2.shp")

# output the polygons layer as a shapefile
#st_write(seg.cls, outname, append = FALSE)

```


