---
title: "Dates"
author: "Rosemary Hartman"
date: "2/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = F)
```
# Dates

Dealing with dates is one of the most difficult data types, but fortunately the package "lubridate" makes it much easier.

## Date and Time classes

There are a few different types of data that reffer to dates. "Dates" just have a date component, whereas date-times may be POSIXct (number of seconds since 1970), or PSIXlt (a named vector of year/month/day/hour/minute/second/time zone).



```{r}
library(tidyverse)
library(lubridate)
?lubridate

?DateTimeClasses

today()

now()
rightnow = now()

rightnow

class(today())
class(now())

```

## Converting from characters to dates

I made a little example spreadsheet with dates and times in different formats. The basic "read.csv" function will automatically read these in as character vectors.

```{r}

dates = read.csv("dates.csv", stringsAsFactors = F)
str(dates)

```
It thinks "Date" is a vector of characters. We want to tell it that it's a date.

In base R, we can use as.date()

```{r}

dates2 = as.Date(dates$Date1)

#Oops

?as.Date
?strptime


#if we don't tell it the format we are in trouble.
dates2 = as.Date(dates$Date1, format = "%m/%d/%Y")

str(zoopdates)

```



Well, that seemed to work, but I can never remember how the "format" thing works. Fortunately, lubridate has some shortcuts. 



```{r}
?ymd

dates2 = mutate(dates, Date1 = mdy(Date1))

str(dates2)

#there are similar functions to reformat the other date structures

dates2 = mutate(dates, Date1 = mdy(Date1), #mdy automatically converts "month/day/year" or "month-day-year" 
                Datetime = mdy_hm(Datetime), #mdy_hm converts both dates and times stored together (usually the best way to do it)
                Date2 = ymd(Date2),
                Date3 = mdy(Date3))

str(dates2)

```

## Times

R has a lot of ways of dealing with dates and times, and it can handle dates without times, but it actually has no built-in way of dealing with times that don't have dates. For that we can use the "hms" package (included in the "tidyverse")

```{r}
dates3 = mutate(dates2, StartTime = hms::parse_hm(StartTime),
                EndTime = hms::parse_hm(EndTime)) #parse_hm converts it to a time

str(dates3)

```
OK, so that worked pretty well, but you notice that "1955" got a little messed up. R stores date/times as the number of seconds since 1970, so if you feed R a 2-diget year, it assumes anything below 70 should be prefaced with "20" and anything above 70 should be prefaced with "19". This is why you should never store 2-diget years!

## Extracting parts of a date-time

If I want to organize my dataset by month, I need to extract the "month" part of the date.

```{r}

?month

month(today())

month(today(), label = TRUE)

dates3 = mutate(dates3, Month = month(Date1), 
                 julianday = yday(Date1), 
                 monthday = mday(Date1), 
                 weekday = wday(Date1))
View(dates3)
```

## Arithmatic with dates and times

Lubridate can store periods of time in three formats 

* Period - Change in clock time between two date-times (years, months, days, etc)
* Interval - Period between two specific time points (from 12:42 today until 2:16 tomorrow)
* Duration - The exact number of seconds in a time span
* difftime - base R version of a timespan

```{r}

rightnow - now()

timetest = rightnow - now()

class(timetest)

as.period(timetest)


span = interval(ymd_hms("2009-01-01 00:00:00"), ymd_hms("2010-02-02 01:01:01")) #interval
as.period(span)
as.period(span, unit = "day")


as.interval(timetest, start = now())

as.duration(timetest)
as.duration(span)

```

# Now you try it!

1. What happens if you parse a string that contains invalid dates?

```{r}

ymd(c("2010-10-10", "cheese"))

```
2. Parse each of the following dates (use lubridate or base r, as you see fit)

```{r}
d1 <- "January 1, 2010"

d2 <- "2015-Mar-07"

d3 <- "06-Jun-2017"

d4 <- c("August 19 (2015)", "July 1 (2015)")

d5 <- "12/30/14" # Dec 30, 2014

```

3. How many seconds are in the month of March?


4. How old are you in minutes?




