---
title: "Loops in R"
author: "Travis Hinkelman"
date: "2019-11-14"
output: 
  ioslides_presentation:
    css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Idiomatic R

>Of course someone has to write loops. It doesn't have to be you. - [Jenny Bryan](https://speakerdeck.com/jennybc/row-oriented-workflows-in-r-with-the-tidyverse?slide=16)

But, if you are like me, you might find loops *more* intuitive for some tasks.

And, in some cases, a loop is the best tool for the job.

## Sequence

Hand-written loops are more verbose and less optimized than built-in functions.

```{r echo = TRUE}
x <- c()
val <- 1
while (val < 2.1) {
  x <- c(x, val)
  val <- val + 0.1
}
x
seq(1, 2, 0.1)
```

## Repeat

```{r echo = TRUE}
x <- c()
for (i in 1:10){
  x <- c(x, "*")
}
x
rep("*", 10)
```

## Print

```{r echo = TRUE}
for (i in 1:5){
  x <- c()
  for (j in 1:i){
    x <- c(x, "*")
  }
  cat(c(x, "\n"), sep = "")
}
```

## Print

```{r echo = TRUE, eval = FALSE}
for (i in 1:5){
  x <- c()
  for (j in 1:i){
    x <- c(x, "*")
  }
  cat(c(x, "\n"), sep = "")
}
```

```{r echo = TRUE}
for (i in 1:5){
  cat(c(rep("*", i), "\n"), sep = "")
}
```

## Fill Arrays

Classic use case for loops.

```{r echo = TRUE}
out <- array(dim = c(3, 3, 3))
for (i in 1:3){
  for (j in 1:3){
    for (k in 1:3){
      out[i, j, k] <- i * j + k
    }
  }
}
out[, , 1]
```

## Manipulate Lists

First, we need to create a list for the next loop example.

```{r echo = TRUE}
rlist <- list(A = data.frame(x = sample(10, 3), y = sample(10, 3)),
              B = data.frame(x = sample(10, 3), y = sample(10, 3)))
rlist
```

## Manipulate Lists

If creating a new column in every data frame in the list, a novice R programmer might start with the following: 

```{r echo = TRUE}
rlist[["A"]]$z <- rlist[["A"]]$x + rlist[["A"]]$y
rlist[["B"]]$z <- rlist[["B"]]$x + rlist[["B"]]$y
```

Reasonable choice for one-off script when list has few elements.

## Manipulate Lists

However, if list has many elements,

```{r echo = TRUE, eval = FALSE}
rlist[["A"]]$z <- rlist[["A"]]$x + rlist[["A"]]$y
rlist[["B"]]$z <- rlist[["B"]]$x + rlist[["B"]]$y
rlist[["C"]]$z <- rlist[["C"]]$x + rlist[["A"]]$y
rlist[["D"]]$z <- rlist[["D"]]$x + rlist[["D"]]$y
rlist[["E"]]$z <- rlist[["E"]]$x + rlist[["E"]]$y
rlist[["F"]]$z <- rlist[["F"]]$x + rlist[["F"]]$y
```

then it is shorter and less error prone to write a loop.

```{r echo = TRUE}
for (i in names(rlist)){
  rlist[[i]]$z <- rlist[[i]]$x + rlist[[i]]$y
}
```

## Manipulate Lists

```{r echo = TRUE}
for (i in names(rlist)){
  rlist[[i]]$z <- rlist[[i]]$x + rlist[[i]]$y
}
```

Same loop written with `mutate` (from [`dplyr`](https://dplyr.tidyverse.org) package) to add the column.

```{r echo = TRUE, eval = FALSE}
for (i in names(rlist)){
  rlist[[i]] <- dplyr::mutate(rlist[[i]], z = x + y)
}
```

## Manipulate Lists

```{r echo = TRUE}
for (i in names(rlist)){
  rlist[[i]]$z <- rlist[[i]]$x + rlist[[i]]$y
}
```

Same loop written with `mutate` (from [`dplyr`](https://dplyr.tidyverse.org) package) to add the column.

```{r echo = TRUE}
for (i in names(rlist)){
  rlist[[i]] <- dplyr::mutate(rlist[[i]], z = x + y)
}
```

Pairing `lapply` and an anonymous function allows you to skip the loop.

```{r echo = TRUE}
rlist <- lapply(rlist, function(x){x$z = x$x + x$y; x})
```

## Generate Data

Crude approach to simulate acoustic telemetry data of emigrating juvenile salmon.

```{r echo = TRUE}
sample5 <- function(x){sample(x, 5, replace = TRUE)}

out_list <- list()
for (i in 1:20){
  out_list[[i]] <- 
    data.frame(day = sort(runif(20, 0, 50)),
               river_mile = c(sample5(20:16),
                              sample5(15:11),
                              sample5(10:6),
                              sample5(5:1)))
}
out_df <- dplyr::bind_rows(out_list, .id = "cohort_id")
```

## Generate Data

```{r echo = TRUE}
head(out_df, 10)
```

## Save Plots

```{r echo = TRUE, eval = FALSE}
library(ggplot2)
for (i in unique(out_df$cohort_id)){
  out_sub <- dplyr::filter(out_df, cohort_id == i)
  ggplot(out_sub, aes(x = day, y = river_mile)) +
    geom_line() +
    geom_point() +
    labs(title = paste("Cohort", i))
  ggsave(filename = paste0("Cohort_", i, ".png"), 
         width = 6, 
         height = 4, 
         path = "figures")
}
```

## Write Files

Splitting data frames
```{r echo = TRUE, eval = FALSE}
for (i in unique(out_df$cohort_id)){
  out_sub <- dplyr::filter(out_df, cohort_id == i)
  write.csv(x = out_sub,
            file = file.path("output", paste0("Cohort_", i, ".csv")),
            row.names = FALSE)
}
```

Elements of a list
```{r echo = TRUE, eval = FALSE}
for (i in names(out_list)){
  write.csv(x = out_list[[i]],
            file = file.path("output", paste0("Cohort_", i, ".csv")),
            row.names = FALSE)
}
```

## Read Files

```{r echo = TRUE, eval = FALSE}
fn <- list.files(path = "output", pattern = "csv", full.names = TRUE)

in_list <- list()
for (i in seq_along(fn)){
  in_list[[i]] <- read.csv(fn[i])
}
```

Skip the loop with `lapply`.

```{r echo = TRUE, eval = FALSE}
in_list <- lapply(fn, read.csv)
```

## Split-Apply-Combine

```{r echo = TRUE}
out_list <- list()
for (i in unique(PlantGrowth$group)){
  pg_sub <- dplyr::filter(PlantGrowth, group == i)
  avg_weight_sub <- mean(pg_sub$weight)
  out_list[[i]] <- data.frame(avg_weight = avg_weight_sub)
}
dplyr::bind_rows(out_list, .id = "group")
```

## Split-Apply-Combine

```{r echo = TRUE, eval = FALSE}
out_list <- list()
for (i in unique(PlantGrowth$group)){
  pg_sub <- dplyr::filter(PlantGrowth, group == i)
  avg_weight_sub <- mean(pg_sub$weight)
  out_list[[i]] <- data.frame(avg_weight = avg_weight_sub)
}
dplyr::bind_rows(out_list, .id = "group")
```

Skip the loop with `group_by` and `summarise` from the `dplyr` package.

```{r echo = TRUE, eval = FALSE}
PlantGrowth %>% 
  group_by(group) %>% 
  summarise(avg_weight = mean(weight))
```

## Take-away Points

When available, use vectorized functions.

Loops are useful when filling arrays, reading/writing data, and saving plots.

Loops are natural extension of procedural style of programming.

Growing a data structure in a loop is inefficient.  
Pre-allocate instead.

If you don't find loops intuitive, explore the [`apply`](https://nsaunders.wordpress.com/2010/08/20/a-brief-introduction-to-apply-in-r/) family of functions or the [`purrr`](https://purrr.tidyverse.org) package.

## Take-away Points

>I was just reminded I've gone 782 days with no loops written in #rstats. Why do we need to teach these to beginners again? - [Miles McBain](https://twitter.com/MilesMcBain/status/966849825634369536)

